---
title: "Spatial Intro 03: Data About Data -- Intro to Metadata File Formats and Structure"
date:   2015-10-27
authors: [Leah A. Wasser]
contributors: [Megan Jones ]
dateCreated: 2016-09-27
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [raster, rgdal, eml, devtools]
category: [self-paced-tutorial]
tags: [R, spatial-data-gis, metadata]
mainTag: spatial-data-management-series
workshopSeries: [spatial-data-management-series]
description: "This tutorial covers metadata stored in a structured,
 Ecological Metadata Language (EML) format."
code1: /R/dc-spatio-temporal-intro/03-structured-metadata-EML.R
image:
  feature:
  credit:
  creditlink:
permalink: /R/structured-metadata-EML
comments: true
sidebar:
  nav: "spatial-intro"
author_profile: false
---


This tutorial covers what metadata - Ecological Metadata Language (EML).

**R Skill Level:** Beginner

<div class="notice--success" markdown="1">


# Goals / Objectives

After completing this activity, you will:

* Understand that metadata come in many formats and must be maintained with the
data.
* Understand that there is necessary metadata associated with and/or embedded in
the data.
* Understand that collecting data requires and includes collecting the metadata.
* Be able to create metadata files for different data types.


### Install R Packages


**OPTIONAL** This tutorial uses these `R` packages in the examples.
 If you want to follow along, please make sure the following packages
 are installed:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`
* **devtools** `install.packages("devtools")`
* NOTE: You have to call the devtools library `library(devtools)` first, and then `install_github` will work.
* **eml** `install_github("ropensci/EML", build=FALSE, dependencies=c("DEPENDS", "IMPORTS"))`

* [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

****


### Additional Resources


* Information on the
<a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank"> `raster` R package</a>

</div>



## Structured Metadata - EML

The Ecological Metadata Language (EML) is a data specification developed
specifically to document ecological data. An EML file is created using a **XML-
based format**.

This means that content is embedded within hierarchical tags. For example
the title of a dataset might be embedded in a `<title>` tag as follows:

	 <title>Fisher Meteorological Station at Harvard Forest since 2001</title>

Similarly, the creator of a dataset is also be found in a hierarchical tag
structure.

    <creator>
      <individualName>
        <givenName>Emery</givenName>
        <surName>Boose</surName>
      </individualName>
    </creator>


## EML Terminology

Let's first discuss some basic **EML** terminology. In the
context of EML, a file documents a **dataset**. This dataset may consist of one
or more files that are documented in the EML document. In the case of our
tabular meteorology data, the structure of our EML file includes:

1. The **dataset**: A dataset may contain one or more data tables associated
with it that may contain different types of related information. This
Harvard meteorological dataset contains data tables with the measurements
collected at the tower.
2. The **data tables**: Data tables refer to the actual data that make up the
dataset. For the Harvard dataset, each data table contains a suite of
meteorological metrics including precipitation and temperature (and associated
quality flags) that are aggregated at a particular time interval (e.g. one data
table contains monthly average data, another contains 15 minute averaged data,
etc)


## Work With EML in R

The `EML` package for `R` is designed to open, read and create EML formatted
metadata. In this tutorial, we will demonstrate how we can use EML
structured metadata in an automated workflow.

NOTE: To save time, we will not explicitly teach the `EML` package given it is
still being developed. But we will provide an example of how you can access
EML metadata programmatically using the `EML` package.

To begin, we will load the `EML` package directly from
<a href="https://github.com/ropensci" target="_blank">ropensci's Git repository </a>.

```{r install-EML-package, results="hide", warning=FALSE }
# install R EML tools
library("devtools")
install_github("ropensci/EML", build=FALSE, dependencies=c("DEPENDS", "IMPORTS"))

# load ROpenSci EML package
library("EML")

# load ggmap for mapping
library(ggmap)


# EML / data location
# http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001
# table 4 http://harvardforest.fas.harvard.edu/data/p00/hf001/hf001-04-monthly-m.csv
```

Next, we will read in the Harvard Forest LTER `EML` file directly from the
online URL using the `read_eml()` function. This file describes multiple data
products that are available for downloaded on the
<a href="http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001" target="_blank"> Harvard Forest Data Archive Page for Fisher Meteorological Station.</a>

Note that because this EML file is large, it takes quite a few seconds for the file to
load.


```{r read-eml, warning=FALSE, message=FALSE}
# import EML from Harvard Forest Met Data

# note: the original xml file is below commented out
# eml_HARV <- read_eml("http://harvardforest.fas.harvard.edu/data/eml/hf001.xml")
# import a truncated version of the eml file for quicker demonstration
eml_HARV <- read_eml("http://neon-workwithdata.github.io/NEON-R-Spatio-Temporal-Data-and-Management-Intro/hf001-revised.xml")

# view size of object
object.size(eml_HARV)

# view the object class
class(eml_HARV)

```

The `read_eml()` function creates an `EML` class object. This object can be
accessed using `slots` in `R` (`@`) rather than a typical subset `[ ]` approach.

## Explore Metadata Attributes

We can begin to explore the contents of our EML file and associated data that it
describes. Let's start at the **dataset** level. We can use `slots` to view
the contact information for the dataset and a description of the methods.

```{r view-eml-content }
# view the contact name listed in the file

eml_HARV@dataset@creator

# view information about the methods used to collect the data as described in EML
eml_HARV@dataset@methods


```


## Identify & Map Data Location

Looking at the coverage for our data, there is only one unique x and y value.
This suggests that our data were collected at one point (x, y) location. We
know this is data from a tower so a point location makes sense. Let's grab the
x and y coordinates and create a quick context map. We will use `ggmap` to create our map.

<i class="fa fa-star"></i> **Data Note:**  If this were a rectangular extent
(e.g. from a raster, line or polygon object) we'd want the bounding box, not
just a point within the extent. We need the extent to properly geolocate and
process the data.
{: .notice}


```{r map-location, warning=FALSE, message=FALSE}

# grab x coordinate from the coverage information
XCoord <- eml_HARV@dataset@coverage@geographicCoverage[[1]]@boundingCoordinates@westBoundingCoordinate@.Data

# grab y coordinate from the coverage information
YCoord <- eml_HARV@dataset@coverage@geographicCoverage[[1]]@boundingCoordinates@northBoundingCoordinate@.Data

# map <- get_map(location='Harvard', maptype="terrain")

# plot the NW corner of the site.
map <- get_map(location='massachusetts', maptype="toner", zoom=8)

ggmap(map, extent=TRUE) +
  geom_point(aes(x=as.numeric(XCoord), y=as.numeric(YCoord)),
             color="darkred", size=6, pch=18)

```

The above example demonstrates how we can extract information from an **EML**
document and use it programmatically in `R`! This is just the beginning of what
we can do!

* Learn More:<a href="https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf" target="_blank"> A nice cheatsheet for GGMAP created by NCEAS</a>

## Unstructured Metadata - Web pages & Text Files

Some metadata are stored in a less or non-structured format such as a web page
or a text file.

Let's visit the
<a href="http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001" target="_blank"> Harvard Forest Fisher Tower webpage</a>
to explore some unstructured metadata.

Visiting this page, we see a list of data files available for Harvard Forest
that can be downloaded directly. Hovering over the files with our mouse, we can
see that these are .csv tabular text files.

Scroll down to the **Overview** section. Here we start to see information that
can be considered metadata about the data available for download.


<div id="challenge" markdown="1">

## Challenge - Explore Unstructured Metadata

Explore the metadata stored on the Harvard Forest LTER web page. Answer the
following questions.

1. What is the time span of the data available for this site?
2. You have some questions about these data. Who is the lead investigator / who
do you contact for more information? And how do you contact them?
3. Where is this field site located? How is the site location information stored
in the metadata? Is there any information that may be useful to you viewing the
location? (HINT: what if you were not familiar with Harvard as a site or are
from another country, etc?)
4. Field Site Information: What is the elevation for the site? What is the
dominant vegetation cover for the site? HINT: Is dominant vegetation easy to
find in the metadata?
5. How is snow dealt with in the precipitation data?
6. Are there some metadata attributes that might be useful to access in a script
in `R` or `Python` rather than viewed on a web page?

HINT: Can you answer all of the questions above from the information provided
on this website? Is there information that you might prefer to find on that page?
</div>

```{r challenge-code-metadata, echo=FALSE}
# Metadata Notes from hf001_10-15-m_Metadata.txt
# 1. 2001-2015
# 2. Emery Boos - located at the top of the document, email is available
# 3. a lat long is available in the metadata at the top, we see the location # described as Prospect Hill Tract (Harvard Forest).
# 4. 342 m elevation, the veg type is not clear in the metadata
# 5. Found in the methods: Delayed melting of snow and ice (caused by problems with rain gage heater or heavy precipitation) is noted in log - daily values are corrected if necessary but 15-minute values are not. The gauge may underestimate actual precipitation under windy or cold conditions.
# 6. this could be a discussion. things like units, time zone, etc are all useful
# if accessed programmatically

```

NOTE: This data and metadata are used in the
[*Introduction to Working With Time Series Data in Text Formats in R* series](http://www.neondataskills.org/tutorial-series/tabular-time-series/).


## Unstructured vs. Structured Metadata

Metadata are particularly important when we are working with data that we did
not collect ourselves. When you download, or gain access to data from a
colleague, or other data provider, be sure to first find and review the
metadata. Then, if the data is in a seperate file (e.g., structured or
unstructured metadata) make sure you save the metadata in a directory that is
closely associated with wherever you save the data itself!


## Create Metadata For Your Data

When you are creating data that you want to share with others, it is critical
to create your own metadata. While this is beyond the scope of this tutorial,
There are many resources available to support metadata documentation including:

* <a href="https://knb.ecoinformatics.org/#tools/morpho" target="_blank"> **Morpho**</a>:
an open-source program that can be used to enter metadataa to be stored in a
file that conforms to the Ecological Metadata Language (EML) specifications.

<div id="challenge" markdown="1">
## Challenge: Find & Explore Metadata

Refer back to the list of possible data sources at the end of
[*Answer a Spatio-temporal Research Question with Data: Where to Start?*]({{ site.baseurl}}/R/spatio-temporal-research-questions#resources-for-locating-spatial-and-ecological-data).
Choose a site that is relevant to your interests and then find a dataset. For
this data set find the metadata and answer the following questions:

1. What format are the metadata in?
2. What metric is collected? Does it have units? If so, what?
3. Who collected the data? Using what type of instruments?

</div>

## ATBD vs Metadata

An Algorithm Theoretical Basis Document (ATBD) is a document which describes the
collection and processing methods associated with a data product. While these
documents are beyond the scope of this tutorial, it is important to note that
often important information about your data may be documented in documents
such as ATBDs.

* <a href="http://data.neonscience.org/documents" target="_blank">NEON</a>
provides ATBDs for all of it's data products.
